Archaic introgression and the distribution of shared functional variation under stabilizing selection

### 2023-11-21 doc/manuscript.tex
start fleshig out discussion

$F_2$ at a given locus depends on the demographic history
relating the two populations and the effect size at the locus due to selection
on the trait. In the infinitesimal limit, involving many loci each of
vanishingly small effect, dynamics at a given locus will be approximately
neutral, so that $F_2$ depends only on the demography. In this case,
\begin{align}
    V_G & \approx f V_{G,0} + (1-f) V_{G,1} + 2f(1-f)\mathbb{E}[F_2] \sum_l a_l^2 \\
    \nonumber
    & = f V_{G,0} + (1-f) V_{G,1} + 2f(1-f)\mathbb{E}[F_2] L V_M,
\end{align}
where $L$ is the number of trait affecting loci.

@@

\begin{itemize}
    \item Stabilizing selection with a constant optimum is an appropriate
        null model for the dynamics of trait variation and selection on
        alleles underlying those traits
    \item Stabilizing selection results in population genetic predictions
        for patterns of introgressed variation in and around functional
        regions that differ from both load- and incompatibility-based arguments
        \begin{itemize}
            \item Load-based models predict that haplotypes from the less ``loaded''
                population will be favored under introgression in either direction
                -- thus, regions of depleted ancestry after introgression in one
                direction will not be expected to also be introgressed ancestry
                deserts after admixture in the opposite direction
            \item \citet{mueller1942} originally pointed out that hybrid
                incompatibilities should not form at single loci -- classical
                incompatibilities require two or more loci, which are not
                expected to be tightly linked to one another. Thus, under a
                model of hybrid incompatibilities, the minor parent ancestry
                is again selected against, but those loci will be in different
                regions. Again, deserts are therefore not expected to overlap
                if selection is largely due to incompatibilities
            \item Instead, stabilizing selection causes introgressed DNA to be
                depleted at the same loci under bi-directional gene flow
                (could be thought of as single-locus incompatibility effects --
                but these are not incompatibilities in the classical sense,
                as the same effects occur even without admixture)
            \item A prediction we could make is that, if current models are correct
                that there has been introgression in both directions between
                \emph{H. sapiens} and Neanderthals, introgressed ancestry deserts
                should overlap with each other more often than expected by chance
            \item See Figure~\ref{fig:ancestry-deserts}
        \end{itemize}
    \item Shared deserts are most likely to form where moderate- or large-effect
        alleles have fixed in one population or the other -- this could provide
        an approach to map\dots?
    \item Selection in structured populations with ongoing migration or
        meta-populations -- SFS approach can handle this -- prospects for the
        future
    \item Some conjectures about hybrid zones
    \item Pleiotropy\dots!!!
\end{itemize}

@@

>>>>

$F_2$ at a given locus depends on the demographic history
relating the two populations and the effect size at the locus due to selection
on the trait. In the infinitesimal limit, involving many loci each of
vanishingly small effect, dynamics at a given locus will be approximately
neutral, so that $F_2$ depends only on the demography. In this case,
\begin{align} \label{eq:VG-admixture}
    V_G & \approx f V_{G,0} + (1-f) V_{G,1} + 2f(1-f)\mathbb{E}[F_2] \sum_l a_l^2 \\
    \nonumber
    & = f V_{G,0} + (1-f) V_{G,1} + 2f(1-f)\mathbb{E}[F_2] L V_M,
\end{align}
where $L$ is the number of trait affecting loci.

@@

Many phenotypic traits are under stabilizing selection
\citep{hodgins2015gene,sanjak2018evidence,sella2019thinking}. This motivates
stabilizing selection around a shared optimum as an appropriate null models for
the dynamics of allele frequencies affecting polygenic traits
\citep{yair2022population}. Stabilizing selection results in distinct dynamics
for trait-affecting alleles, characterized by symmetric underdominance
\citep{robertson1956effect,keightley1988quantitative}, that is, selection
against the minor allele. Because admixture and introgression events are common
in natural populations, it will be important to understand the effects of gene
flow on the architecture of complex traits.  \aprcomment{or\dots``studies of
complex traits in admixed populations rely on this understanding, or
something''}

After admixture between diverged populations, the additive genetic variance of
a trait rapidly increases. The genetic variance after admixture depends on both
the existing genetic variances within the parental population and their
divergence at trait-affecting loci, measured by $F_2$
(Equation~\ref{eq:VG-admixture}). $V_G$ is then expected to fairly quickly
decay to background levels. During this process, introgressed and pre-existing
trait-affecting alleles are replaced by new mutations, turning over the genetic
architecture of the trait (Figure\ref{fig:toy-admixture}). Selection occurs
against the minor alleles at trait-affecting loci, so that introgressed alleles
from the minor parental ancestry, whether derived or reintroduced ancestral
alleles, are selected against (Figure~\ref{fig:linkage-pred}).

Selection against introgressed alleles will also remove linked introgressed
ancestry in the surrounding regions. Thus, introgressed ancestry deserts are
more likely to form at and around loci contributing to selected traits
(Figure~\ref{fig:deserts}). This process is symmetric, so that deserts tend to
form in the same regions under reciprocal introgression. When comparing
distributions of introgressed haplotypes in two diverged populations with more
recent gene flow, we should expect to see an overlap in deserts of introgressed
ancestry at loci affecting traits under stabilizing selection.

The expected pattern of shared ancestry deserts under stabilizing selection
differs from models of both deleterious load and incompatibilities. Load-based
models predict that haplotypes that have accumulated more deterious mutations,
e.g., from a population with small long-term effective population size, will be
selected against under either direction of gene flow. Introgressed ancestry at
a given selected locus will decrease in frequency in one introgression scenario
and increase the other. This may explain the replacement of MT and Y chromosome
DNA in Neanderthals by human haplotypes after early human-to-Neanderthal
introgression \citep{posth2017deeply,petr2020evolutionary}.

\citet{muller1942isolating} originally hypothesized that hybrid
incompatibilities should not often form within a single locus, but instead the
two or more loci involved will most often be found at distant or unlinked loci
in genome. Theory predicts that hybrid incompatibilities are resolved via
selection against the minor parental ancestry [cite]. Depending on the
direction of introgression, different incompatibilities alleles will be
selected against and ancestry deserts will form around those different loci.
While such a model predicts selection against introgressed ancestry, similar to
the stabilizing selection model, an important distinction is that deserts are
not expected to overlap with bidirectional gene flow.

\aprcomment{any data about distributions of incompatibility loci in the
genome?}

\aprcomment{Second-to-last paragraph}
Turning to observations from human and Neanderthal genomes
\citep{harris2023diverse}

\aprcomment{last paragraph}
\begin{itemize}
    \item Selection in structured populations with ongoing migration or
        meta-populations -- SFS approach can handle this -- prospects for the
        future
\end{itemize}
\aprcomment{end on a comment about the usefulness of the SFS approach and the
promise for future studies of dynamics of the architecture of complex traits
with non-trivial demographic scenarios -- possible that this should be earlier,
and we can include caveats there. Maybe second paragraph of discussion.}

\aprcomment{don't end on this, or might not need to include it -- caveats can
come earlier?}
\begin{itemize}
    \item Some conjectures about hybrid zones - single-locus ``incompatibilities''
    \item Pleiotropy\dots!!!
\end{itemize}

@@

\section*{Acknowledgements}
\begin{itemize}
    \item Lab members
    \item Kevin Thornton
    \item Bret Payseur
\end{itemize}
